---
title: "Indian household clustering"
author: "Luke Wolcott"
date: "May 28, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

### The question

Is there a natural way to cluster Indian households, based on data about the nature of these households -- do they have electricity, a TV, a computer, a cow, an ox-drawn cart?  How do these clusters correspond to the 5-level Wealth Index?

### The dataset

The [National Family Health Survey](http://rchiips.org/nfhs/data1.shtml) is conducted in India every 10 years or so.  The latest one was in 2015-2016 and the data is yet to be released.  In preparation, I'm looking at the third NFHS, conducted in 2005-2006.  I registered to get access to this data, (which unfortunately I can't upload to GitHub).

This week, I'm looking at how Indian households cluster according to their characteristics.    The dataset includes 109,041 separate questionnaire responses, each with 3588 (!!!) variables entered.  A codebook can be found in the file `IAHR52FL.MAP` (open it with a text editor).  There's a copy of the household questionnaire starting on p.636 of `FRIND3-Vol1AndVol2.pdf`. These files are uploaded to the GitHub repo corresponding to this report.

### The variables

I will look specifically at the following variables.   

| Variable | Does the household have...?   | Variable | [other characteristics] |
|:---------|:-----------------------------|:---------|:----|
| HV206    | electricity                 | HV025    | Rural or urban? |
| HV207     |     radio                   | HV013 | # in household? |
| HV208 | television               | HV014 | # of children in HH? |
| HV209 | refrigerator            | HV216 | # of rooms for sleeping? |
| HV210 | bicycle                  | HV219     | sex of HH head? |
| HV211 | motorcycle/scooter       | HV220 | age of HH head? |
| HV212 | car                     | SH30 | anyone with tuberculosis? | 
|HV221 | non-mobile telephone       |SH58 | own this house? |
| HV227 | bednets for sleeping |    SH62A | have cows/bulls/buffalo? |
| HV243A | mobile telephone |       SH62B | have camels? |
| HV243B | watch |                  SH62C | have horses/donkeys/mules? |
| HV243C | animal-drawn cart |      SH62D | have goats? |
| HV246 | livestock |               SH62E | have sheep? |
| SH47B | mattress |                SH62F | have chickens? |
| SH47C | pressure cooker |         SHSTRUC | nuclear or non-nuclear family? |
| SH47D | chair |
| SH47E | cot/bed |
| SH47F | table |
| SH47G | electric fan |
| SH47I | B&W television |
| SH47J | color television |
| SH47K | sewing machine |
| SH47N | computer |
| SH47U | water pump |
| SH47V | thresher |
| SH47W | tractor |
| SH56A | any windows |
| SH56B | windows with glass |
| SH56C | windows with screens |
| SH56D | windows with curtains/shutters |


### TDA to detect clusters

```{r}
library(dplyr)

# the original dataset was cleaned using the code in clean.R,
# run on a sample of 1000 cases from the 109041 total.
d <- read.csv("thousand_cases.csv", header=TRUE)

# subset for the columns we're interested in
df <- select(d, HV025, HV206,HV207,HV208,HV209,HV210,HV211,HV212,HV219,HV221,
             HV227,HV243A,HV243B,HV243C,HV246,SH30,SH47B,SH47C,SH47D,
             SH47E,SH47F,SH47G,SH47I,SH47J,SH47K,SH47N,SH47U,SH47V,SH47W,
             SH56A,SH56B,SH56C,SH56D,SH58,SH62A,SH62B,SH62C,SH62D,SH62E,SH62F,
             SHSTRUC,
             HV013,HV014,HV216,HV220)

# if factor variables, convert to 0/1.  
# otherwise, center at zero and scale so sd=1.
for (i in 1:ncol(df)){
      if (is.factor(df[,i]))
            df[,i] <- as.numeric(df[,i])-1
      else
            df[,i] <- scale(as.numeric(df[,i]))
}

# there are a handful of rows with some NAs, and this will take them out
df <- df[!is.na(apply(df,1,sum)),]

# now df is 993 x 44
```


As a first step, we can use topological data analysis (TDA) to see if there is a natural number of clusters in the data.  For the sake of computation, I'll sample 1000 of the 109,041 cases (and remove 7 with missing data).  To do the whole 993 x 44 dataset with maxscale=10, it takes my laptop about 7.5 minutes.  It only took about 2 minutes with maxscale=5.  Below we have the persistence homology barcode and the persistence diagram.

(See [this report](https://lukewolcott.github.io/WaterAndRainInLakeSuperior/water_and_rain.html) or [this report](https://lukewolcott.github.io/TDAwithSimulatedData/TDAwithSimulatedData.html) for some details on using TDA.)


```{r, cache=TRUE}
library(TDA)
maxscale <- 10
maxdimension <- 1

# compute the Rips complex
Diag <- ripsDiag(X = df, maxdimension, maxscale, library = "GUDHI", location=TRUE)

# plot persistence barcode
plot(Diag[["diagram"]], barcode = TRUE)
mtext("Barcode of 0D and 1D persistent homology")
legend("right", lty=c(1,1), lwd=c(3,3), col=c("red", "black"), legend=c("H1", "H0"))

# plot persistence diagram
plot(Diag[["diagram"]], barcode = FALSE)
mtext("Persistence diagram of 0D and 1D persistent homology")
legend("right", lty=c(1,1), lwd=c(3,3), col=c("red", "black"), legend=c("H1", "H0"))
```


Because there are 993 black barcode lines, it's hard to see how many larger clusters we are looking at.  For this we can look explicitly at the birth and death of the H0 intervals.

```{r}
# output 15 longest persistence intervals
Diag[["diagram"]][1:15,]
```

We see that before everything gets clumped together into one connected component, there is a long stretch with two clusters.  And there is a persistent feature of k=5 clusters.  So, miraculously or by design, it was a good idea to use a Wealth Index with values 1-5.

### What are the two clusters?

Before we cluster with k=5, let's see what we can say about the situation with k=2, which was suggested by TDA.  It seems the two clusters have similar sizes.

```{r}
set.seed(134)

# cluster into 2 clusters, output table of assignments
km2 <- kmeans(df,2,nstart=20)
table(km2$cluster)
```

We can tabulate the mean for each variable, grouped by cluster, to look for any large differences.

```{r}
options(digits=3)
df2 <- cbind(df,cluster2=km2$cluster)
x <- df2 %>% group_by(cluster2) %>% summarise_all(mean)
as.data.frame(x)
options(digits=8)
```

If we plot histograms of a few of these variables, the picture emerges that cluster 1 tends to be rural households, with almost no amenities like a refrigerator or cell phone.  Cluster 2 tends to be urban households, almost all electrified, and much more likely to have household amenities.


```{r}
library(ggplot2)

# use gridExtra library to make 2x2 plots here
library(gridExtra)

g <- ggplot(df, aes(fill=as.factor(km2$cluster))) +       
      theme(legend.title=element_blank())
 
p1 <- g + geom_bar(aes(HV206), position = "dodge") + 
      labs(x="HV206: electricity")
p2 <- g + geom_bar(aes(HV209), position = "dodge") + 
      labs(x="HV209: refrigerator")
p3 <- g + geom_bar(aes(HV243A), position = "dodge") + 
      labs(x="HV243A: cell phone")
p4 <- g + geom_bar(aes(HV025), position = "dodge") + 
      labs(x="HV025: rural(0) or urban(1)")

grid.arrange(p4,p1,p3,p2,ncol=2,nrow=2,top=quote("Comparing two clusters"))
```


### Comparing clusters by Wealth Index

Now let's return to the full dataset of 109,041 households, and use the same 44 variables to cluster with k=5.  

The supplemental file `NFHS3SUP.pdf`, p.6, explains how the "Wealth Index" variable HV270 was constructed using PCA to generate weight for these variables.  The index is a cumulative score, normalized to a ranking from 1 (poorest) to 5 (wealthiest).  How do the five clusters compare with the Wealth Index?

```{r, cache=TRUE}
# all_cases_cleaned.csv is the full IAHR52FL.SAV cleaned by clean.R
df_full <- read.csv("all_cases_cleaned.csv", header=TRUE)

# cluster into 5 clusters, ignoring the Wealth Index column
# larger nstart throws a memory error on my laptop
km5 <- kmeans(df_full[,1:45],5,nstart=3) 

table(km5$cluster, df_full$HV270)
```

Cluster 5 seems to capture most of the Richest and Richer households.  Cluster 4 is capturing a lot of the Richer, with some bleeding into Richest and Middle.  Cluster 3 is roughly evenly spread between Middle, Poorer, and Poorest.  Cluster 2 and 1 do not bin well with the Wealth Index, although cluster 2 has few Richest households and cluster 1 has few Poorest ones.

```{r}
g2 <- ggplot(df_full, aes(fill=as.factor(km5$cluster)))
g2 + geom_bar(aes(HV270), position = "dodge")
```



---
