---
title: "‘Women’s March crowds’ vs Non-Trump Votes"
author: "Luke Wolcott"
date: "January 30, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

On Saturday January 21, 2017, the day after Trump's inauguration, thousands of people protested around the US (and around the world).  Previously I looked at cities with the largest protest crowds, adjusted by city population.  Now I'll compare statewide protest crowd totals to the election vote tallies.

Specifically, I ask the question: **What is the relationship between a state's 1/21/17 protest total and the number of people who voted against Trump?**

### Crowd number data

Again, I'll use data gathered by Jeremy Pressman and Erica Chenoweth, taken from this site on 1/24/2017:

`https://docs.google.com/spreadsheets/d/1xa0iLqYKz8x9Yc_rfhtmSOJQ2EGgeUVjvV4A8LsIaxY/htmlview?sle=true#`

I clean up this data a little, take out non-US protests, and group protests by state.  

```{r}
# spreadsheet was downloaded from above link, then cleaned up in Excel a little.  The non-US data was removed.

# this code loads the data and cleans it up more
data <- read.csv("USCrowdNumbers.csv", stringsAsFactors = FALSE, na.strings="0")
names(data) <- c("City", "StateCode", "Estimate1", "Estimate2")
data$Estimate1 <- sub(',','', data$Estimate1)
data$Estimate2 <- sub(',','', data$Estimate2)
data$Estimate1 <- as.numeric(data$Estimate1)
data$Estimate2 <- as.numeric(data$Estimate2)
data <- data[!is.na(data$Estimate1),]

# create column for average of low and high crowd estimates
data$CrowdAverage <- (data$Estimate1 + data$Estimate2)/2

# group by StateCode, using dplyr.  
library(dplyr)
by_state <- group_by(data, StateCode)
CrowdByState <- summarize(by_state, StateCrowd = sum(CrowdAverage))
CrowdByState <- as.data.frame(CrowdByState)
bad <- c("--", "CA/NV", "VI", "GM")
CrowdByState <- CrowdByState[!is.element(CrowdByState[,1], bad),]
```

### Election tallies

Our 2017 National Election data comes from David Wasserman and the Cook Political Report, compiled here:

`https://docs.google.com/spreadsheets/d/133Eb4qQmOxNvtesw2hdVns073R68EZx4SfCnP4IGQf8/htmlview?sle=true#gid=19`

I combine  the votes for "Clinton" and "Others" into "NonTrump", organized by state.  

```{r}
votes <- read.csv("NonTrumpVotesClean.csv", stringsAsFactors = FALSE)
```

### State protest crowds versus Non-Trump election votes

```{r}
data <- merge(votes, CrowdByState)
data$Ratio <- data$StateCrowd/data$NonTrump
#hist(data$Ratio, breaks=15)
```

Washington, DC, is an outlier because many people came from out of the city.  The ratio of protesters to population is 1.976!

Here is a plot of StateCrowd versus Non-Trump Votes, on a log10 scale, which allows you to see which state is where by hovering.

```{r}
library(plotly)
plot_ly(data, x=~log10(NonTrump), y=~log10(StateCrowd), mode="markers", text = ~StateCode)
```

We are interested in fitting a line to the data.

```{r}
x <- log10(data$NonTrump)
y <- log10(data$StateCrowd)
fit <- lm(y ~ x)
plot(x,y, pch=20, xlab = "Non-Trump votes (log scale)", ylab = "Protest crowd (log scale)", main = "State protest size vs Non-Trump votes")
abline(fit, lwd=2, col="blue")
```

The fitted slope is 0.9375, which means that if State B has ten times as many non-Trump voters as State A, we would expect the crowd to be 10^(0.9375) = 8.66 times larger.

A non-log plot also shows the correlation.
```{r}
xx <- data$NonTrump
yy <- data$StateCrowd
fitt <- lm(yy ~ xx)
plot(xx,yy, pch=20, xlab="Non-Trump votes", ylab="Protest crowd", main="State protest size vs Non-Trump votes")
abline(fitt, lwd=2, col="blue")
```

The slope for this fitted line is 0.06997, which we can interpret as saying that on average 7.00% of non-Trump voters attended the 1/21/2017 protest.  (Except that the crowd estimates include non-voters, for example children.)

If we remove DC, we can more reasonably compare the 50 states.  Here is a heatmap showing each state's ratio of protest crowd to non-Trump votes.  


```{r}
data_noDC <- data[-8,]
names(data_noDC)[1] = "State"
data_noDC$hover <- with(data_noDC, paste(State))#, '<br>',"Ratio:",Ratio))
borders <- list(color = toRGB("red"))
map_options <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)
plot_ly(data_noDC, z = ~Ratio, text = ~hover, locations = ~State, type = 'choropleth', locationmode = 'USA-states', color = ~Ratio, colors = 'Blues', marker = list(line = borders)) %>% layout(title = 'Ratio of protest crowd to non-Trump votes', geo = map_options)
```

