---
title: "Art Grants in LA"
author: "Luke Wolcott"
date: "March 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

---

In mid-January The Hill [reported](http://thehill.com/policy/finance/314991-trump-team-prepares-dramatic-cuts) that Trump was considering cutting funding to the National Endowment for the Arts.  The Heritage Foundation, a conservative think tank that was [reportedly very influential](http://www.cnn.com/2016/12/06/politics/donald-trump-heritage-foundation-transition/) in Trump's transition process, [argues that](http://www.huffingtonpost.com/entry/trump-reportedly-plans-to-end-national-arts-funding_us_5880de61e4b070d8cad0e184) 

> “Taxpayer assistance of the arts is neither necessary nor prudent [...] Taxpayers should not be forced to pay for plays, paintings, pageants, and scholarly journals, regardless of the works’ attraction or merit.”

I thought it would be interesting to look at data on public funding of the arts, and I found an interesting dataset for Los Angeles County.  The question is: **what types of art are funded, and how much?**

---

The dataset ([downloaded from here](https://data.lacounty.gov/Arts-and-Culture/Organizational-Grants-for-Arts-Services/73tm-r9gy)) contains information about art grants given through the Organizational Grant Program of the Los Angeles County Arts Commission, over the three application years of 2014-15, 2015-16, and 2016-17.  

These plots were done with the plotly library, so they have some nice interactivity to them.


```{r}
# Data and cleaning
data <- read.csv("Organizational_Grants_for_Arts_Services.csv")
library(plotly)
library(dplyr)
```


```{r}
# Convert the grant award amount to a numeric that is easy to work with.  
# Fix a typo in the name of one of the disciplines.
data$Grant.Award <- sub('.','',as.character(data$Grant.Award))
data$Grant.Award <- as.numeric(data$Grant.Award)
for (i in 1:568)
      if (data[i,2] == "Arts Service/")
            data[i,2] <- as.factor("Arts Service ")
```


### A. Spread of money awarded, by discipline

```{r, echo=FALSE}
plot_ly(data, y = ~Grant.Award, color = ~Discipline, type = "box")
```

### B. Spread of money awarded, by application year

```{r, echo=FALSE}
plot_ly(data, y = ~Grant.Award, color = ~Application.Year, type="box")
```

### C. Number of grants awarded versus total amount awarded, by discipline

```{r, echo=FALSE}
by_disc <- group_by(data, Discipline)
d <-summarize(by_disc, Total.Amount.Awarded = sum(Grant.Award), Number.Awarded = length(Grant.Award))
plot_ly(d, x=~Number.Awarded, y=~Total.Amount.Awarded, mode="markers", text = ~Discipline)
```

---
